An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
V2 after revisions in response to reviewer reports

```{r load-packages, include=FALSE, cache=FALSE}
#Load the packages needed to read and work with the data
library(grid)
library(gridExtra)
library(kableExtra)
library(png)
library(sf)
library(tidyverse)
library(units)
library(vaccHamilton) # Data package
```

@@

```{r save-population-map, include=FALSE}
ggplot() +
  geom_sf(data = data_da_2016 %>%
            filter(`Region Name` == "Hamilton") %>%
            dplyr::select(Population_55to69, Population_70plus) %>%
            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
            mutate(Age = factor(Age,
                                levels = c("Population_55to69",
                                           "Population_70plus"),
                                labels = c("Age: 55 to 69",
                                           "Age: 70+"),
                                ordered = TRUE)) %>%
            st_as_sf(),
          aes(fill = Population),
          color = NA) +
  geom_sf(data = urban_types,
          aes(color = Type),
          fill = NA) +
  scale_fill_distiller(name = "Population",
                       breaks = c(20, 50, 150, 400),
                       palette = "OrRd", 
                       trans = "log",
                       direction = 1) +
  scale_color_manual(name = "",
                     values = c("Urban" = "firebrick2",
                                "Suburban" = "cornflowerblue",
                                "Rural" = "forestgreen"))  +
  facet_wrap(~ Age, nrow = 2) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))

@@

```{r save-figure-maps-baseline, include=FALSE}
jpeg(filename = "images/baseline-maps.jpeg")
grid.arrange(ttm_map_baseline +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline  +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
dev.off()

ttm_map_baseline
ggsave("images/ttm-baseline-map.jpeg")

pht_map_baseline
ggsave("images/pht-baseline-map.jpeg")
```

@@

```{r save-figure-results, include=FALSE}
jpeg(filename = "images/figure-results.jpeg")
grid.arrange(ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_income + 
               ggtitle("By income: Scenario") + 
               labs(x = "", y = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             ttm_region_baseline + 
               ggtitle("By region: Pilot Program") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_region + 
               ggtitle("By region: Scenario") + 
               labs(y = "") +
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             nrow = 2)
dev.off()

ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40))
ggsave("images/ttm-income-pilot.jpeg")

ttm_region_baseline + 
  ggtitle("By region: Pilot Program") + 
  xlim(c(0,40))
ggsave("images/ttm-region-pilot.jpeg")

ttm_income +
  ggtitle("By income: Scenario") + 
  labs(x = "", y = "") + 
  xlim(c(0,40))
ggsave("images/ttm-income.jpeg")

ttm_region + 
  ggtitle("By region: Scenario") + 
  labs(y = "") +
  xlim(c(0,40))
ggsave("images/ttm-region.jpeg")
```

>>>>

```{r load-packages, include=FALSE, cache=FALSE}
#Load the packages needed to read and work with the data
library(grid)
library(gridExtra)
library(kableExtra)
library(patchwork)
library(png)
library(sf)
library(tidyverse)
library(units)
library(vaccHamilton) # Data package
```

@@

```{r save-population-map, include=FALSE, fig.height=8, dpi=300}
ggplot() +
  geom_sf(data = data_da_2016 %>%
            filter(`Region Name` == "Hamilton") %>%
            dplyr::select(Population_55to69, Population_70plus) %>%
            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
            mutate(Age = factor(Age,
                                levels = c("Population_55to69",
                                           "Population_70plus"),
                                labels = c("Age: 55 to 69",
                                           "Age: 70+"),
                                ordered = TRUE)) %>%
            st_as_sf(),
          aes(fill = Population),
          color = NA) +
  geom_sf(data = urban_types,
          aes(color = Type),
          fill = NA) +
  scale_fill_distiller(name = "Population",
                       breaks = c(20, 50, 150, 400),
                       palette = "OrRd", 
                       trans = "log",
                       direction = 1) +
  scale_color_manual(name = "",
                     values = c("Urban" = "firebrick2",
                                "Suburban" = "cornflowerblue",
                                "Rural" = "forestgreen"))  +
  facet_wrap(~ Age, nrow = 2) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))

@@

```{r save-figure-maps-baseline, include=FALSE, fig.height=8, dpi=600}
ttm_plot <- ttm_map_baseline +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7))

pht_plot <- pht_map_baseline  +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7))

ttm_plot/pht_plot

ggsave("images/baseline-maps.jpeg")
```

@@

```{r save-figure-results, include=FALSE, fig.height=8, dpi=900}
ttm_income_baseline <- ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none")

ttm_income <- ttm_income + 
               ggtitle("By income: Scenario") + 
               labs(x = "", y = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7))

ttm_region_baseline <- ttm_region_baseline + 
               ggtitle("By region: Pilot Program") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none")

ttm_region <- ttm_region + 
               ggtitle("By region: Scenario") + 
               labs(y = "") +
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7))

(ttm_income_baseline + ttm_income) / (ttm_region_baseline + ttm_region)

ggsave("images/figure-results.jpeg")
```

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
V2 after revisions in response to reviewer reports

Along with the provision of health care facilities to treat severe cases
of COVID-19 (e.g., Ghorbanzadeh et al., 2021; Pereira et al., 2021a),
another front in the fight against the pandemic is the rolling out of
vaccination programs. The Government of Ontario, in Canada, implemented
in 2021 a pilot program to offer vaccines in pharmacies for people 60
years and older. In the first incarnation of this program, the province
approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the
latter including the region of Frontenac-Lennox and Addington). The City
of Hamilton, the third most populous urban center in the province,
operated dedicated vaccination centers for people aged 70+ and mobile
pop-up clinics for people aged 75+; however, it did not have any
approved pharmacies under the pilot for other populations until April
1st 2021, with the announcement of 20 pharmacies in the city
\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
At the same time that the expansion of the pilot was announced, the
province extended eligibility for this program to people 55 years and
older.

@@

As seen in Table \ref{tab:distribution-results}, under the pilot program
approximately 36.42\% of people live in DAs in the bottom 40\% of the
median household income scale, but they account for 51.98\% of the total
person-hours of travel. In contrast, 44.5\% of people aged 55 to 69 in
DAs in the top 40\% of the median household income scale accrue only
35.03\% of the total person-hours of travel. Where the mean travel time
of residents of DAs with high median household income is 347 minutes,
residents of lower income DAs average 746 minutes in travel time. In
addition to longer average travel time, residents in lower income DAs
also see substantially larger variations in travel times, and some may
face considerably longer travel times (see top-left panel in Figure
\ref{fig:results}).

@@

For comparison purposes we consider a scenario with some modest
additions to the list of pharmacies in the provincial pilot. For the
scenario we choose three plausible sites (i.e., pharmacies that belong
to chains already used by the province for this program). We do not
conduct a systematic search of candidate locations. We note that some
papers propound the design optimal accessibility landscapes (e.g.,
Horner, 2008; Páez et al., 2013), but such an endeavor is beyond the
scope of a short findings paper. Instead, we repeat the analysis after
including the three urban sites shown in white circles in Figure
\ref{fig:pharmacies-and-regions}.

>>>>

Along with the provision of health care facilities to treat severe cases
of COVID-19 (e.g., Ghorbanzadeh et al., 2021; Pereira et al., 2021a),
another front in the fight against the pandemic is the rolling out of
vaccination programs. The Government of Ontario, in Canada, implemented
a pilot program to offer vaccines in pharmacies for people 60 years and
older in early 2021. In the first incarnation of this program, the
province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston
(the latter including the region of Frontenac-Lennox and Addington). The
City of Hamilton, the third most populous urban center in the province,
operated dedicated vaccination centers for people aged 70+ and mobile
pop-up clinics for people aged 75+; however, it did not have any
approved pharmacies under the pilot for other populations until April
1st 2021, with the announcement of 20 pharmacies in the city
\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
At the same time that the expansion of the pilot was announced, the
province extended eligibility for this program to people 55 years and
older.

@@

As seen in Table \ref{tab:distribution-results}, under the pilot program
approximately 36.42\% of people live in DAs in the bottom 40\% of the
median household income scale, but they account for 51.98\% of the total
person-hours of travel. In contrast, 44.5\% of people aged 55 to 69 in
DAs in the top 40\% of the median household income scale accrue only
35.03\% of the total person-hours of travel. Where the mean travel time
of residents of DAs with high median household income is 347 minutes,
residents of lower income DAs average 746 minutes in travel time. In
addition to greater average travel times, the top-left panel in Figure
\ref{fig:results} suggests residents in lower income DAs also see
substantially larger variability in travel times and some may face
considerably longer trip durations.

@@

For comparison purposes we consider a scenario with some modest
additions to the list of pharmacies in the provincial pilot. For the
scenario we choose three plausible sites (i.e., pharmacies that belong
to chains already used by the province for this program). We do not
conduct a systematic search of candidate locations. We note that some
papers propound the design of optimal accessibility landscapes (e.g.,
Horner, 2008; Páez et al., 2013), but such an endeavor is beyond the
scope of a short findings paper. Instead, we repeat the analysis after
including the three urban sites shown in white circles in Figure
\ref{fig:pharmacies-and-regions}.
